What We Eval in the Shadows

### 2021-04-05 analysis/corpus.Rmd
Add traced evals figure

```{r}
cs_count(PKG_C, evals)
```

@@

```{r}
eval_call_sites_hist <- function(df, df_all, y_lab, max_evals=20L) {
  median_evals <- median(df$evals)
  df %>% 
    mutate(
      evals=ifelse(evals < max_evals, evals, max_evals)
    ) %>%
    ggplot(aes(x=evals)) + 
    geom_histogram(binwidth=1, color="black", fill="blue", alpha=0.3) +
    geom_vline(aes(xintercept=median_evals), linetype="dashed", color="red", size=0.2) +
    scale_x_continuous(
      limits=c(0, max_evals + 1), 
      breaks=as.integer(sort(c(seq(1, max_evals, length.out=4), round(median_evals)))), 
      labels=function(x) c(head(x, -1), str_c(tail(x, 1), "+"))
    ) +
    scale_y_continuous(
      labels=scales::comma_format(accuracy=1),
      sec.axis=sec_axis(
        trans=~./nrow(df_all), 
        labels=scales::percent_format(accuracy=.1), 
        name=y_lab)
    ) +
    labs(y=y_lab) +
    theme(text = element_text(size=21),
          axis.title.x = element_blank(),
          panel.grid.minor.x = element_blank(), aspect.ratio = 4/3) 
}

@@

>>>>

```{r}
cs_count(PKG_C, evals) %>% print(n=Inf)
```

@@

```{r}
eval_call_sites_hist <- function(df, df_all, y_lab, max_evals=20L) {
  median_evals <- median(df$evals)
  df %>% 
    mutate(
      evals=ifelse(evals < max_evals, evals, max_evals)
    ) %>%
    ggplot(aes(x=evals)) + 
    geom_histogram(binwidth=1, color="black", fill="blue", alpha=0.3) +
    geom_vline(aes(xintercept=median_evals), linetype="dashed", color="red", size=0.2) +
    scale_x_continuous(
      limits=c(0, max_evals + 1), 
      breaks=as.integer(sort(c(seq(1, max_evals, length.out=4), median_evals))), 
      labels=function(x) c(head(x, -1), str_c(tail(x, 1), "+"))
    ) +
    scale_y_continuous(
      labels=scales::comma_format(accuracy=1),
      sec.axis=sec_axis(
        trans=~./nrow(df_all), 
        labels=scales::percent_format(accuracy=.1), 
        name=y_lab)
    ) +
    labs(y=y_lab) +
    theme(text = element_text(size=21),
          axis.title.x = element_blank(),
          panel.grid.minor.x = element_blank(), aspect.ratio = 4/3) 
}

@@

### Eval plot

```{r}
eval_plot_df <- PKG_C %>%
  filter(package_r_code > 0) %>%
  select(package, evals, evals_hit, package_r_code)

size_breaks <- c(1e3, 5e3, 10e3, 25e3, 50e3, 100e3)
# size_labels <- str_c(
#   c("≤ ", rep("> ", 6)),
#   scales::comma(accuracy = 1, x=size_breaks[-1])
# )

eval_plot_df %>% 
#  filter(evals > 0, evals_hit > 0) %>%
  mutate(
    # size=cut(package_r_code, size_breaks, size_labels, include.lowest=T),
    label=ifelse(evals %in% tail(sort(evals),3), package, NA)
  ) %>%
  ggplot(aes(x=evals, y=evals_hit, size=package_r_code, label=label)) + 
    geom_jitter(alpha=0.3) +
    geom_text(nudge_y = .15, check_overlap = TRUE, size=4) +
    scale_x_continuous(
      trans="log10",
      breaks=function(x) {
        f <- scales::breaks_log()
        sort(c(f(x), max(eval_plot_df$evals)))
      },
      labels=scales::comma_format(accuracy = 1)
    ) +
    scale_y_continuous(
      trans="pseudo_log",
      breaks=function(x) {
        f <- scales::breaks_log()
        sort(c(0, 1, f(c(1, x[2])), max(eval_plot_df$evals_hit)))
      },
      labels=scales::comma_format(accuracy = 1)
    ) +
    scale_size_binned(
      range=c(1,10),
      breaks = size_breaks,
      labels=scales::comma_format(accuracy = 1)
    ) +
#    geom_smooth(method = "lm", se = FALSE, show.legend = FALSE, linetype="dashed", color="red", size=0.2) +
    labs(x="Eval call sites (log)", y="Traced eval call sites (log)", size="Package code size") +
    theme(legend.position = c(0.15, 0.8))
ggsave(path(PLOT_DIR, "traced-eval-callsites.pdf"))
```

### 2021-04-05 paper/main.tex
Eval provenance
Add traced evals figure

\begin{wrapfigure}{r}{5cm} \hspace*{-12mm}
  \includegraphics[width=75mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
  \eval call sites}%
  \label{fig:pkgs-eval-callsites-hist}
\end{wrapfigure}
%
This section focuses on CRAN packages and reports statistics about the usage of
\eval. There are \PkgEvalCallSites calls to \eval in \PkgPackages packages.
These represent \PkgPackagesRatio of all packages or \PkgCodeRatio of the
entire code size of CRAN. Over half of the packages have less than 3 \eval
call sites and except \MaxEvalCallSitesPackage (\MaxEvalCallSitesCount \eval
calls), all packages contain less than \MaxEvalCallSitesRest \eval call sites
(\cf Fig~\ref{fig:pkgs-eval-callsites-hist}) These \eval calls appear in
\PkgFunsWithEval R functions (\CranFunsWithEvalRatio of all R functions in
CRAN).

For the dynamic analysis, we execute all the \CranRunnableScripts programs
extracted from \CranPackages packages. Any run that did not exercise one of the
\eval call sites in the packages was discarded. This left \Nbruns runs of some
of the \Corpus packages to generate the data presented here. There were
\Allcalls invocations of \eval and its variants occurring in \Triggeredpkgs
packages. Out of \PkgEvalCallSitesRnd \eval call sites, \PkgHitEvalCallSites
were hit at least once\footnote{Not all packages with static call sites of
\eval were observed triggering that call site, this can be chalked down to
incomplete code coverage and occasional tracer failure (\PkgFailedProgramsRatio
of programs do not terminate successfully or within 30 minute timeout).}. On
average, the analysis covers \PkgHitEvalCallSitesAvgRatio of package \eval call
sites, which similar to the \PkgCodeCoverage of average package code coverage.
Figure~\ref{freq} summarizes the frequency of dynamic calls to \eval with the
left column being the number of calls and the right, the number of packages in
that range. There are \Fewcalls packages with low \eval frequency, fewer than
100 calls; and a similar number of packages, \Manycalls to be precise, that use
\eval more than 1,000 times.  One package, \Maxcallspack, calls it \Maxcalls
times and thus accounts for over half of the observed calls.

@@

\mypara{Provenance.}

% Substitute/match.call
% parse/expression
% bytecode?

% Explain tha some generate expression objects and other language objects?
There are various ways to obtain expressions that \eval will evaluate:
\begin{itemize}
    \item \c{substitute} synthesizes ASTs from expressions by replacing
    symbols with their bindings in the specified environment.
    \item \c{match.call} captures a call and its arguments as expressions.
    \item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression.
    \item \c{expression}  creates a vector of expression
    objects from text.
    \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
    expressions.
    \item \c{comp} compiles R code into R bytecode.
\end{itemize}

\begin{table}[h]
  \begin{tabular}{|c|c|c|}
      \hline
      & \#sites & \%sites \\
       \hline
         \c{substitute}, \c{quote} & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\
      \hline
      \c{parse}, \c{str2expression} & \packageNbParseExprsSites & \packageParseExprsSitePercent \\
      \hline
        \c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\   
      \hline
      \c{expression} & \packageNbNotParseExprsSites & \packageNotParseExprsSitePercent  \\
      \hline
      \c{str2lang} & \packageNbStrlangSites  & \packageStrlangSitePercent  \\
      \hline
      \c{comp} & \packageNbBytecodeSites & \packageBytecodeSitePercent \\
      \hline
  \end{tabular}
\caption{Provenances of \eval expressions. }
\end{table}

>>>>

This section focuses on CRAN packages and reports statistics about the usage of
\eval. There are \PkgEvalCallSites calls to \eval in \PkgPackages packages.
These represent \PkgPackagesRatio of all packages or \PkgCodeRatio of the
entire code size of CRAN. Over half of the packages have less than 3 \eval
call sites and except \MaxEvalCallSitesPackage (\MaxEvalCallSitesCount \eval
calls), all packages contain less than \MaxEvalCallSitesRest \eval call sites
(\cf Fig~\ref{fig:pkgs-eval-callsites-hist}) These \eval calls appear in
\PkgFunsWithEval R functions (\CranFunsWithEvalRatio of all R functions in
CRAN).

For the dynamic analysis, we execute all the \CranRunnableScripts programs
extracted from \CranPackages packages. Any run that did not exercise one of the
\eval call sites in the packages was discarded. This left \Nbruns runs of some
of the \Corpus packages to generate the data presented here. There were
\Allcalls invocations of \eval and its variants occurring in \Triggeredpkgs
packages. Out of \PkgEvalCallSitesRnd \eval call sites, \PkgHitEvalCallSites
were hit at least once (\cf Fig~\ref{fig:traced-eval-callsites}). On average,
the analysis covers \PkgHitEvalCallSitesAvgRatio of package \eval call sites,
which similar to the \PkgCodeCoverage of average package code
coverage\footnote{Not all packages with static call sites of \eval were
observed triggering that call site, this can be chalked down to incomplete code
coverage and occasional tracer failure (\PkgFailedProgramsRatio of programs do
not terminate successfully or within 30 minute timeout).}.  Figure~\ref{freq}
summarizes the frequency of dynamic calls to \eval with the left column being
the number of calls and the right, the number of packages in that range. There
are \Fewcalls packages with low \eval frequency, fewer than 100 calls; and a
similar number of packages, \Manycalls to be precise, that use \eval more than
1,000 times.  One package, \Maxcallspack, calls it \Maxcalls times and thus
accounts for over half of the observed calls.

\begin{figure}[t]
\begin{minipage}{.3\textwidth}
  \vspace*{14mm}
  \hspace*{-10mm}
  \includegraphics[width=1.5\textwidth]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
  \eval call sites}%
  \label{fig:pkgs-eval-callsites-hist}
\end{minipage}
\begin{minipage}{.6\textwidth}
  \centering
  \includegraphics[width=\textwidth]{traced-eval-callsites.pdf}
  \caption{\eval call sites coverage of the \PkgPackages packages.}%
  \label{fig:traced-eval-callsites}
\end{minipage}
\end{figure}

@@

\paragraph{Provenance.} 
% Explain tha some generate expression objects and other language objects?
There are various ways to obtain expressions that \eval will evaluate:
\begin{itemize}
    \item \c{substitute} synthesizes ASTs from expressions by replacing
    symbols with their bindings in the specified environment.
    \item \c{match.call} captures a call and its arguments as expressions.
    \item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression.
    \item \c{expression}  creates a vector of expression
    objects from text.
    \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
    expressions.
    \item \c{comp} compiles R code into R bytecode.
\end{itemize}

\begin{table}[h]
  \begin{tabular}{|c|c|c|}
      \hline
      & \#sites & \%sites \\
       \hline
         \c{substitute}, \c{quote}, ... & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\
      \hline
      \c{parse}, \c{str2expression} & \packageNbParseExprsSites & \packageParseExprsSitePercent \\
      \hline
        \c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\   
      \hline
      \c{expression} & \packageNbNotParseExprsSites & \packageNotParseExprsSitePercent  \\
      \hline
      \c{str2lang} & \packageNbStrlangSites  & \packageStrlangSitePercent  \\
      \hline
      \c{comp} & \packageNbBytecodeSites & \packageBytecodeSitePercent \\
      \hline
  \end{tabular}
\caption{Provenances of \eval expressions. }
\end{table}
